DOpElib: A Goal Oriented Software Library for Computing PDEs and Optimization Problems

### 2012-10-05 Paper_May_2012/main.tex
Paper: Nonstationary problems hinzugefuegt und mit Optimization problems begonnen und Ideen fuer dessen Realisierung im Paper hinzugefuegt

\author{CHRISTIAN GOLL
\affil{University of Heidelberg}
THOMAS WICK
\affil{University of Heidelberg}
WINNIFRIED WOLLNER
\affil{University of Hamburg}}

@@

\begin{bottomstuff}
Author's adresses: C. Goll {and} T. Wick, Institut f\"ur Angewandte Mathematik, Universit\"at Heidelberg;
W. Wollner, Fachbereich Mathematik, Universit\"at Hamburg.
\end{bottomstuff}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Detailed Description of the Main Features}
\label{detailed_description}
\todo{Aus der Sicht des Users anhand eines konkreten Beispiels beschreiben}
\todo{Einfaches Beispiel nehmen}
\todo{erwaehnen, dass wir alles immer nicht-linear betrachten}
\todo{Aussehen eines typischen PDE und Opt loop und deren Gemeinsamkeiten
erklaeren. Vielleicht mir Pseudocode}
This library is designed to allow easy implementation and numerical solutions 
of problems involving partial differential equations (PDEs). 
\subsection{Solving of PDEs}
\subsubsection{Stationary problems}
The easiest case 
is that of a PDE in weak form to find some $u$
\[
a(u)(\phi) = 0 \quad \forall \phi \in V,
\]
with some appropriate space $V$. As example, we take 
the Laplace problem in weak form:
\[
a(u)(\phi) = (\nabla u, \nabla \phi) \quad \forall \phi \in V.
\]
To solve the problem, we need a configuration (a mesh), eventually 
a diffusion parameter, a finite element, a quadrature rule,
a linear solver (here the CG solver), and some output such as
a graphical solution and eventually some measurement functional.
Each of those routines are written in the \texttt{main.cc} file. 
We start by including necessary files both the DOpElib library and
the deal.II software:
\begin{lstlisting}
#include "pdeproblemcontainer.h" etc. ...
\end{lstlisting}
In the \texttt{main} function, the following steps need to be done. First, 
the parameter reader for runtime parameters is initialized:
\begin{lstlisting}
  ParameterReader pr;
  SSolver::declare_params(pr);
  DOpEOutputHandler<VECTOR>::declare_params(pr);
  pr.read_parameters(paramfile);
\end{lstlisting}
A linear finite element in two dimensions and two components is 
initialized:
\begin{lstlisting}
  FESystem<2> state_fe(FE_Q<2>(1), 2);
\end{lstlisting}
and a quadrature rule of order three is used to approximate the 
integrals:
\begin{lstlisting}
  QGauss<2> quadrature_formula(3);
\end{lstlisting}
Then, the \texttt{LocalPDE} object comes into play. This is an quite 
important feature of DOpElib because it allows for easy implementation 
of the variational formulation. Specific explication is given below because
the implementation is seperated in \texttt{localpde.h}.
Afterwards, a \texttt{LocalPointFunctional} is declared. As for the PDE, the 
implementation is done in another file, namely \texttt{localfunctional.h}.

Generically, each problem in DOpElib is considered to be time-dependent and nonlinear because
these are the 'real' world problems. Therefore, (even for the linear and stationary Laplace equation),
we have to declare a (pseudo) time:
\begin{lstlisting}
  std::vector<double> times(1, 0.);
\end{lstlisting}
Then, the spatial mesh (here the unit square) is created and three-times globally-refined:
\begin{lstlisting}
  GridGenerator::hyper_cube(triangulation, 0, 1);
  triangulation.refine_global(3);
\end{lstlisting}
A major component which organizes the whole solution process (independent from 
the problem!) is the 
\begin{lstlisting}
MethodOfLines_StateSpaceTimeHandler<FE, DOFHANDLER, SPARSITYPATTERN, VECTOR, 2> DOFH(
      triangulation, state_fe);
\end{lstlisting}
Then, the problem (here the Laplace PDE) is given to the \texttt{DOFH} via
\begin{lstlisting}
  OP P(LPDE, DOFH);
\end{lstlisting}
It remains to declare boundary conditions for the problem under consideration. 
Let's simply take zero Dirichlet conditions on the boundary:
\begin{lstlisting}
  std::vector<bool> comp_mask(2);

@@

        for (unsigned int i = 0; i < n_dofs_per_cell; i++)
        {
          for (unsigned int j = 0; j < n_dofs_per_cell; j++)
          {
            local_entry_matrix(i, j) += scalar_product(phi_grads_u[j], phi_grads_u[i]) 
                * state_fe_values.JxW(q_point);
          }
        }
      }
\end{lstlisting}
The cost functional is now implemented in \todo{hier morgen weiter machen}

\subsubsection{Nonstationary problems}
In this section, we focus on the extension to 
the numerical solution of nonstationary processes. 
As example, we consider the heat equation:

...

@@

\begin{lstlisting}
 void CellTimeEquation(const CellDataContainer<dealii::DoFHandler<dealdim>, VECTOR, dealdim>& cdc,
                           dealii::Vector<double> &local_cell_vector __attribute__((unused)),
		       double scale __attribute__((unused)))
    {
// implement \partial_t u in weak form
}
\end{lstlisting}
and it is done in a straight-forward way:
\begin{lstlisting}
 for (unsigned int q_point=0;q_point<n_q_points;q_point++)
	{
	  Tensor<1,2> v;
	  u[0] = _uvalues[q_point](0);
	  u[1] = _uvalues[q_point](1);

@@

	      local_cell_vector(i) +=  scale * (u * phi_i_u) * state_fe_values.JxW(q_point);
	    }
	}
\end{lstlisting}

\subsection{Solving of optimization problems}
More complex cases involve optimization problems given in the form (OPT)
\begin{align*}
\min\;&J(q,u) \\
  &\text{s.t.}\; a(q,u)(\phi) = 0 \quad \forall \phi\in V,\\
  &a \le q \le b,\\
  &g(q,u) \le 0,  
\end{align*}
where $u$ is a FE-function and $q$ can either be a FE-function or some 
fixed number of parameters, $a$ and $b$ are constraint bounds for the control $q$,
and $g(\cdot)$ is some state constraint.

@@

%\subsubsection{Time stepping schemes}
%\todo{Kurz halten}
%Third, at least for nonstationary PDEs we need to select a time stepping scheme
%the file names of which are mostly self explanatory:
%\item \texttt{include/forward\underline{ }euler\underline{ }problem.h}
%\item \texttt{include/shifted\underline{ }crank\underline{ }nicolson\underline{ }problem.h}
%\item \texttt{include/backward\underline{ }euler\underline{ }problem.h}
%\item \texttt{include/fractional\underline{ }step\underline{ }theta\underline{ }problem.h} Note that the use of this scheme requires a special Newton solver, which is, however, already
%implemented for the convenience of the user!
%\item \texttt{include/crank\underline{ }nicolson\underline{ }problem.h}
%\subsubsection{Integrator routines}
%Finally, we need to select a way how to integrate and solve linear and nonlinear equations
%\item \texttt{templates/integrator.h} This class computes integrals over a given 
%  triangulation (including its faces).
%\item \texttt{templates/integrator\underline{ }multimesh.h} The same as above but it is 
%  possible that some of the FE functions are defined on an other triangulation 
%  as long as the have a common coarse triangulation.
%\item \texttt{templates/integratormixeddims.h} This is used to compute integral which 
%  are given in another (larger) dimension than the current variable. (This is exclusively
%  used if the control variable is given by some parameters. Which means \texttt{dopedim == 0}). 
%\subsubsection{Nonlinear solvers}
%\item \texttt{templates/newtonsolver.h} This solves some nonlinear equation using a 
%  line-search Newton method.
%\item \texttt{templates/newtonsolvermixeddims.h} The same but in the case when there is 
%  another variable in a (larger) dimension is involved. See 
%  \texttt{integratormixeddims.h}.
%\item \texttt{templates/instat\underline{ }step\underline{ }newtonsolver.h} This is a 
%  Newton method as above to invert the next time-step. It differs from the plain vanilla
%  version in that it computes certain data from the previous time step only once 
%  and not in every Newton iteration.
%\item \texttt{templates/fractional\underline{ }step\underline{ }theta\underline{ }step\underline{ }newtonsolver.h} This is the Newton solver for the time step in a 
%  fractional-step-theta scheme. It combines the computation of all three sub steps.
%\subsubsection{Linear solvers}
%To solve the linear equations within Newton's method, we use 
%standard solvers from iterative type 
%such as the method of the conjugate gradients (CG method), 
%GMRES, or direct solvers such as \texttt{UMFPACK}.
%In addition, we DOpE provides  a wrapper class 
%\texttt{templates/voidlinearsolver.h} for certain cases when we 
%  know that the matrix to be inverted is the identity. It simply copies the rhs to the
%  lhs. This is only needed for compatibility reasons some other components.
%%\begin{itemize}
%%\item \texttt{templates/cglinearsolver.h} This is a wrapper for the cg solver implemented in 
%%  \texttt{deal.II}. The solver will build and store the stiffness matrix for the PDE.
%%\item \texttt{templates/gmreslinearsolver.h} This is a wrapper for the gmres solver 
%%  implemented in \texttt{deal.II}. The solver will build and store the stiffness matrix 
%%  for the PDE.
%%\item \texttt{templates/directlinearsolver.h} This is a wrapper for the direct solver 
%%  implemented in \texttt{deal.II} using \texttt{UMFPACK}. 
%%  The solver will build and store the stiffness matrix for the PDE.
%%\item \texttt{templates/voidlinearsolver.h} This is a wrapper for certain cases when we 
%%  know that the matrix to be inverted is the identity. It simply copies the rhs to the
%%  lhs. This is only needed for compatibility reasons some other components.
%%\end{itemize}
%%\subsection{Problem specific classes}
%%The following classes are used to describe the problem and will usually require 
%%some implementation.
%%\begin{itemize}
%%  \item \texttt{basic/constraints.h} This is used by the spacetimehandlers to 
%%    compute the number of constraints from the control and state vectors. 
%%    It must not be reimplemented by the user, but needs to be properly 
%%    initialized if OPT is used with box control constraints or $g(q,u) \le 0$.
%%  \item \texttt{interfaces/constraintinterface.h} This gives an interface for both 
%%    the control box constraints as well as the general constraint $g \le 0$. This 
%%    needs to be specified if constraints are to be used. If they are not needed 
%%    a default class \texttt{problemdata/noconstraints.h} can be used. We note that we only 
%%    need to write the integrands on 
%%    elements or faces the loop over elements will be taken care of in the integrator.
%%  \item \texttt{interfaces/pdeinterface.h} This defines an interface for the 
%%    partial differential equation $a(q,u)(\phi) = 0$. This needs to be written
%%    by the user. We note that we only need to write the integrands on 
%%    elements or faces the loop over elements will be taken care of in the integrator.
%%    Specifically, derivatives are written therein, too.
%%  \item \texttt{interfaces/dirichletdatainterface.h} This gives an interface to the 
%%    Dirichlet data for a problem. If the Dirichlet data are simply a function 
%%    (and do not depend on the control $q$) one can use the default class\\
%%    \texttt{problemdata/simpledirichletdata.h}.
%%\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions and Outlook}
\label{conclusions}
In this article, we described the features 
and applications of the DOpElib project, which 
was initiated at the Heidelberg University in 2010.
Specifically, the applications cover a broad 
spectrum of numerical exmamples motivated 
by different disciplines. 
Currently, we are developing concepts for the efficient 
numerical solution of time-dependent optimization
problems governed by PDEs. 
\todo{Was wollen wir noch in den nächsten 1-2 Jahren erreichen?}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgment}
The second author thanks Rolf Rannacher from
the Institute for Applied mathematics at the 
Heidelberg University for financial suppport.

>>>>

\author{CHRISTIAN GOLL
\affil{University of Heidelberg}
THOMAS WICK
\affil{The University of Texas at Austin}
WINNIFRIED WOLLNER
\affil{University of Hamburg}}

@@

\begin{bottomstuff}
Author's adresses: C. Goll, Institut f\"ur Angewandte Mathematik,
Universit\"at Heidelberg;
T. Wick, The Institute for Computational Engineering and Sciences;
W. Wollner, Fachbereich Mathematik, Universit\"at Hamburg.
\end{bottomstuff}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Detailed Description of the Main Features}
\label{detailed_description}
%\todo{Aus der Sicht des Users anhand eines konkreten Beispiels beschreiben}
%\todo{Einfaches Beispiel nehmen}
%\todo{erwaehnen, dass wir alles immer nicht-linear betrachten}
%\todo{Aussehen eines typischen PDE und Opt loop und deren Gemeinsamkeiten
%erklaeren. Vielleicht mir Pseudocode}
This library is designed to allow easy implementation and numerical solutions 
of problems involving partial differential equations (PDEs). 
\subsection{Solving of PDEs}
\subsubsection{Stationary problems}
The easiest case 
is that of a PDE in weak form to find some $u$
\[
a(u)(\phi) = 0 \quad \forall \phi \in V,
\]
with some appropriate space $V$. As example, we take 
the Laplace problem in weak form:
\[
a(u)(\phi) = (\nabla u, \nabla \phi) \quad \forall \phi \in V.
\]
To solve the problem, we need a configuration (a mesh), eventually 
a diffusion parameter, a finite element, a quadrature rule,
a linear solver (here the CG solver), and some output such as
a graphical solution and eventually some measurement functional.
Each of those routines are written in the \texttt{main.cc} file. 
We start by including necessary files both the DOpElib library and
the deal.II software:
\begin{lstlisting}
#include "pdeproblemcontainer.h" etc. ...
\end{lstlisting}
In the \texttt{main} function, the following steps need to be done. First, 
the parameter reader for runtime parameters is initialized:
\begin{lstlisting}
  ParameterReader pr;
  SSolver::declare_params(pr);
  DOpEOutputHandler<VECTOR>::declare_params(pr);
  pr.read_parameters(paramfile);
\end{lstlisting}
A linear finite element in two dimensions and two components is 
initialized:
\begin{lstlisting}
  FESystem<2> state_fe(FE_Q<2>(1), 2);
\end{lstlisting}
and a quadrature rule of order three is used to approximate the 
integrals:
\begin{lstlisting}
  QGauss<2> quadrature_formula(3);
\end{lstlisting}
Then, the \texttt{LocalPDE} object comes into play. This is an quite 
important feature of DOpElib because it allows for easy implementation 
of the variational formulation. Specific explication is given below because
the implementation is seperated in \texttt{localpde.h}.
Afterwards, a \texttt{LocalPointFunctional} is declared. As for the PDE, the 
implementation is done in another file, namely \texttt{localfunctional.h}.

Generically, each problem in DOpElib is considered to be time-dependent and nonlinear because
these are the 'real' world problems. Therefore, (even for the linear and stationary Laplace equation),
we have to declare a (pseudo) time:
\begin{lstlisting}
  std::vector<double> times(1, 0.);
\end{lstlisting}
Then, the spatial mesh (here the unit square) is created and three-times globally-refined:
\begin{lstlisting}
  GridGenerator::hyper_cube(triangulation, 0, 1);
  triangulation.refine_global(3);
\end{lstlisting}
A major component which organizes the whole solution process (independent from 
the problem!) is the 
\begin{lstlisting}
MethodOfLines_StateSpaceTimeHandler<FE, DOFHANDLER, SPARSITYPATTERN, VECTOR, 2> DOFH(
      triangulation, state_fe);
\end{lstlisting}
This method handles all dofs in space and time and different versions exist
for forward as well as optimization problems.

Afterwards, the problem (here the Laplace PDE) is given to the \texttt{DOFH} via
\begin{lstlisting}
  OP P(LPDE, DOFH);
\end{lstlisting}
It remains to declare boundary conditions for the problem under consideration. 
Let's simply take zero Dirichlet conditions on the boundary:
\begin{lstlisting}
  std::vector<bool> comp_mask(2);

@@

        for (unsigned int i = 0; i < n_dofs_per_cell; i++)
        {
          for (unsigned int j = 0; j < n_dofs_per_cell; j++)
          {
            local_entry_matrix(i, j) += scalar_product(phi_grads_u[j], phi_grads_u[i]) 
                * state_fe_values.JxW(q_point);
          }
        }
      }
\end{lstlisting}
The integration of the equation for each DoF and each quadrature point on each
cell is then done 
in the \texttt{templates/integrator.h} function.

The cost functional is now implemented in \todo{hier morgen weiter machen}

\subsubsection{Nonstationary problems}
In this section, we focus on the extension to 
the numerical solution of nonstationary processes. 
As example, we consider the heat equation:
Find $u$:
\[
\partial_t u - \Delta u = 0
\]
and in weak formulation:
Find $u$:
\[
a(u)(\phi) = 0 \quad \forall \phi \in V,
\]
with some appropriate space $V$ and 
\[
a(u)(\phi) = (\partial_t u, \phi) + (\nabla u, \nabla \phi) \quad \forall \phi \in V.
\]
Temporal discretization using the One-step-$\theta$ scheme reads:
Given the old timestep solution $u^n$, find $u^{n+1}$:
\[
\frac{(u^{n+1} - u^{n}, \phi)}{k} + \theta (\nabla u^{n+1}, \nabla \phi)
+ (1 - \theta) (\nabla u^{n}, \nabla \phi) = 0,
\]
Reordering in new and old terms and multiplication by the time step size $k$:
\[
(u^{n+1},\phi) + k \theta (\nabla u^{n+1}, \nabla \phi)
= (u^n, \phi) - (1 - \theta) (\nabla u^{n}, \nabla \phi).
\]
In DOpElib, different time stepping schemes can be found 
in the folder \texttt{DOpEsrc/tsschemes}.

@@

\begin{lstlisting}
 void CellTimeEquation(const CellDataContainer<dealii::DoFHandler<dealdim>, VECTOR, dealdim>& cdc,
                           dealii::Vector<double> &local_cell_vector __attribute__((unused)),
		       double scale __attribute__((unused)))
    {
// implement (u^{n+1},\phi) 
}
\end{lstlisting}
and it is done in a straight-forward way:
\begin{lstlisting}
 for (unsigned int q_point=0;q_point<n_q_points;q_point++)
	{
	  Tensor<1,2> v;
	  u[0] = _uvalues[q_point](0);
	  u[1] = _uvalues[q_point](1);

@@

	      local_cell_vector(i) +=  scale * (u * phi_i_u) * state_fe_values.JxW(q_point);
	    }
	}
\end{lstlisting}
Here, the advantage is that DOpElib orders the terms in the timestepping
problems
such that only $(u^{n+1},\phi)$ must be implemented. However, if someone 
needs explicit input of the old timestep term, then he can use 
\begin{lstlisting}
 void CellTimeEquationExplicit (const CellDataContainer<dealii::DoFHandler<dealdim>, VECTOR, dealdim>& cdc,
                           dealii::Vector<double> &local_cell_vector,
				   double scale)
    {
      // implement (u^{n+1},\phi) and (u^n,\phi)  
    }
    \end{lstlisting}
    For instance, this explicit term is used for fluid-structure interaction in
    which
    the old timestep term needs explicit modification.

\subsection{Solving of optimization problems}
Upgrading the framework for solving optimization problems
is described in this section. A prototypical problem reads:
\begin{align*}
\min\;&J(q,u) \\
  &\text{s.t.}\; a(q,u)(\phi) = 0 \quad \forall \phi\in V,\\
  &a \le q \le b,\\
  &g(q,u) \le 0,  
\end{align*}
where $u$ is a FE-function and $q$ can either be a FE-function or some 
fixed number of parameters, $a$ and $b$ are constraint bounds for the control $q$,
and $g(\cdot)$ is some state constraint.
As before, the main part is done in the \texttt{main} function there is no
great difference to the previous explications. However, 
since we utilize gradient-based optimization routines \cite{...},
all derivatives must be implemented. All this information
must given to the \texttt{localpde.h} file. Herein, one has to 
implement the adjoint, tangent and adjoint Hession equations. 
The algorithmic parts are described in the literature \cite{BeMeVe06}.

Concretely, let us focus on the Laplace equation and ...

this is taken from step-2 in the Opt examples.

\todo{Optimality system}

\todo{Two problems in which we identify the different terms and implementation 
of these terms in DOpE}

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}
\label{conclusions}
In this article, we described the features 
and applications of the DOpElib project, which 
was initiated at the Heidelberg University in 2010.
Specifically, the applications cover a broad 
spectrum of numerical exmamples motivated 
by different disciplines. 
Currently, we are developing concepts for the efficient 
numerical solution of time-dependent optimization
problems governed by PDEs.

\section{Outlook}
\todo{Was wollen wir noch in den nächsten 1-2 Jahren erreichen?}
